% =====================================================================
%  Aethelred — AAAI submission (draft)
%  Compile with the AAAI style files (aaai2026.sty/.bst) in this dir.
%  Figures live in ../../figures ; insight prose mirrors ../aethelred_figures.tex
% =====================================================================
\documentclass[letterpaper]{article}
\usepackage{aaai2026}      % <-- drop the official AAAI style here
\usepackage{times}
\usepackage{helvet}
\usepackage{courier}
\usepackage[hyphens]{url}
\usepackage{graphicx}
\usepackage{xcolor}
\usepackage{booktabs}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{algorithm}
\usepackage{algorithmic}

% ---- figure palette / emphasis macros (match the figures) -----------
\definecolor{aethred}{HTML}{E8112D}
\definecolor{xgnnwine}{HTML}{6A040F}
\definecolor{pgnnor}{HTML}{F77F00}
\definecolor{vinfomag}{HTML}{C9379D}
\definecolor{wingreen}{HTML}{1B7837}
\definecolor{cavorange}{HTML}{B25900}
\newcommand{\aeth}{\textsc{Aethelred}}
\newcommand{\win}[1]{\textcolor{wingreen}{\textbf{#1}}}
\newcommand{\cav}[1]{\textcolor{cavorange}{#1}}

\newtheorem{theorem}{Theorem}
\newtheorem{proposition}{Proposition}
\newcommand{\Ek}{\mathcal{E}_k}
\newcommand{\bigO}{\mathcal{O}}

\pdfinfo{/Title (Aethelred: Deterministic, Tight Certificates for Causally-Faithful GNN Explanations)
         /Author (Anonymous)}
\setcounter{secnumdepth}{2}

\title{Aethelred: Deterministic, Tight Certificates for\\
       Causally-Faithful Graph Neural Network Explanations}
\author{Anonymous Submission}

\begin{document}
\maketitle

% =====================================================================
\begin{abstract}
Explanations of graph neural networks (GNNs) are notoriously fragile: a small,
prediction-preserving perturbation can rewrite the ``reason'' behind a decision.
A recent line of work therefore \emph{certifies} the robustness of an
explanation, guaranteeing that its edges change little under bounded graph
perturbation. We argue that this is not enough: a certificate of
\emph{stability} says nothing about whether the certified subgraph is the
\emph{right} one. A robustly stable explanation of a spurious correlation is
still wrong. We introduce \aeth{}, which certifies the robustness of
\emph{causally faithful} explanations. \aeth{} couples an invariance-trained,
\emph{propagation-free} causal edge scorer with a \emph{deterministic} edge
certificate. Because each edge is scored from its endpoints alone, its
importance is provably invariant to the rest of the graph; this edge-locality
yields a closed-form, per-edge certified radius---no voting and no randomized
smoothing. The resulting certificate is \emph{tight}: a worst-case adversary
achieves exactly the bound. On the certification metric of the state of the art
and its own published numbers, \aeth{} certifies a $3$--$6\times$ larger
perturbation size on real molecular datasets, while a controlled bias sweep
shows its faithfulness is invariant to spurious correlation where prior
explainers collapse. \aeth{} certifies in a single $\bigO(1)$ pass with zero
variance.
\end{abstract}

% =====================================================================
\section{Introduction}
\label{sec:intro}

Graph neural networks (GNNs) drive state-of-the-art results in molecular
property prediction, fraud detection, and scientific discovery, and in all of
these settings an \emph{explanation}---which substructure drove the
prediction---is as important as the prediction itself. Yet GNN explanations are
\emph{fragile}: an adversary can perturb a handful of edges, leave the predicted
label untouched, and completely rewrite the explanatory subgraph. This fragility
undermines exactly the trust that explanations are meant to provide.

A natural response is to \emph{certify} the explanation: to guarantee that, for
any bounded perturbation of the graph, the explanation does not change. The
first certified explainable GNN does this through \emph{randomized-smoothing}
machinery---it hashes the graph into many sub-graphs, explains each, and
\emph{votes}, so that a bounded number of perturbed edges can corrupt only a
bounded number of votes.

\paragraph{Stability without faithfulness is vacuous.}
We observe that certifying stability addresses only half the problem. A
certificate guarantees that the explanation \emph{does not change}; it says
nothing about whether the explanation is \emph{correct}. When a dataset contains
a spurious structural correlation---common in practice and the rule under
distribution shift---a non-causal explainer happily latches onto the spurious
feature, and a stability certificate will then guarantee a \emph{robustly wrong}
explanation. A stable explanation of the wrong subgraph is still wrong. What
should be certified is the robustness of a \emph{causally faithful}
explanation.

\paragraph{This paper.}
We present \aeth{}, a framework that certifies causally-faithful GNN
explanations. \aeth{} has two ingredients. (i) A \emph{causal discovery core}
learns a sparse edge mask via an invariance objective, so the explanation tracks
the causal substructure rather than spurious context. Crucially, the scorer is
\emph{propagation-free}: each edge's importance is a function of its two
endpoints' features alone. (ii) A \emph{deterministic edge certificate} exploits
this edge-locality. Because an edge's score cannot be changed by perturbing any
other edge, the only way an adversary can displace a top-$k$ explanatory edge is
to insert a higher-scoring one; a simple counting argument then yields a
closed-form, per-edge certified radius---with no voting, no smoothing, and zero
variance.

This design gives \aeth{} three properties that prior certified explainers lack.
The certificate is \win{deterministic} (a single forward pass, $\bigO(1)$ rather
than $\bigO(T)$ in the number of voting sub-graphs); it is \win{tight} (a
worst-case adversary achieves exactly the certified bound, leaving no wasted
conservatism, unlike the loose lower bounds of randomized smoothing); and it
certifies a \win{faithful} explanation, so the guarantee is on something that
matters.

\paragraph{Contributions.}
\begin{itemize}
  \item We articulate and empirically demonstrate that \emph{certifying
        explanation stability is insufficient}---a stability certificate can
        guarantee a spurious explanation (Fig.~\ref{fig:thesis}).
  \item We introduce \aeth{}, the first \emph{deterministic} certificate for GNN
        explanations, built on a propagation-free causal scorer whose
        edge-locality we prove yields a closed-form per-edge certified radius
        (Thms.~\ref{thm:overlap}--\ref{thm:hijack}), and which we show is
        \emph{tight} (Fig.~\ref{fig:tight}).
  \item Against the state of the art, on \emph{its own} certification metric and
        published numbers, \aeth{} certifies a $3$--$6\times$ larger
        perturbation size on real molecular data (Fig.~\ref{fig:mlambda}),
        remains faithful under a controlled spurious-correlation sweep where
        prior explainers collapse (Fig.~\ref{fig:thesis}), and certifies at
        $\bigO(1)$ cost with zero variance (Fig.~\ref{fig:eff}).
\end{itemize}

% =====================================================================
\section{Related Work}
\label{sec:related}

\paragraph{GNN explanation and its fragility.}
Post-hoc explainers identify an important subgraph or edge mask for a prediction
\cite{ying2019gnnexplainer,luo2020pgexplainer,wang2021refine,miao2022gsat}. They
optimize predictive sufficiency, not causal correctness, and are known to be
unstable: prediction-preserving perturbations can substantially change the
returned explanation \cite{fragile2024}. This motivates robustness
\emph{guarantees} rather than empirical robustness alone.

\paragraph{Causal / invariant explanation.}
A separate line argues explanations should identify \emph{causal} substructures.
Invariant-rationale methods \cite{wu2022dir} learn a subgraph that remains
predictive across artificial environments \cite{arjovsky2019irm}, discarding
spurious context. These methods improve faithfulness under distribution shift but
provide \emph{no} robustness certificate. \aeth{} inherits the invariance
principle and adds a certificate.

\paragraph{Certified robustness for graphs.}
Certified \emph{prediction} robustness for GNNs is typically obtained by
randomized smoothing / sub-graph voting, yielding probabilistic or deterministic
guarantees on the predicted label under bounded edge or node perturbation.
Recent work extends voting from the label to the \emph{explanation}, certifying
that the explanatory edges overlap the clean (or ground-truth) explanation under
bounded edge perturbation. All of these are \emph{voting} certificates: they
require $T$ sub-graph evaluations, are randomized, and yield conservative
(loose) lower bounds. \aeth{} differs on all three axes---deterministic,
single-pass, and tight---and, distinctively, certifies a \emph{causal}
explanation rather than an arbitrary stable one. Empirical explanation defenses
(e.g.\ information-bottleneck methods) improve robustness without any
certificate; we compare against one as a non-certified reference.

% =====================================================================
\section{Method}
\label{sec:method}

\subsection{Setup and notation}
A graph $G=(V,E)$ has node features $X$. A GNN classifier $f$ predicts a label;
an explanation is a ranking of edges by importance, of which the top-$k$ form the
explanatory subgraph $\Ek$. An adversary may add or delete at most $B$ undirected
edges. We seek to certify that $\Ek$ is stable \emph{and} faithful under any such
$B$-bounded perturbation.

\subsection{Aethelred}
\aeth{} comprises three pillars (Fig.~\ref{fig:arch}).

\paragraph{Causal Discovery Core (propagation-free scorer).}
For each edge $(u,v)$ we compute
\[
  s(u,v)=\sigma\!\big(\mathrm{MLP}[\,h_u,h_v,|h_u-h_v|,\cos(x_u,x_v)\,]\big),
  \quad h_u=\mathrm{MLP}(x_u).
\]
The encoder $h_u$ depends on $x_u$ \emph{only}; no message passing is used. Hence
$s(u,v)$ depends solely on the endpoints of $(u,v)$ and is \emph{invariant to the
presence or absence of every other edge}---the property that makes the
certificate deterministic (Fig.~\ref{fig:locality}).

\paragraph{Focal Engine.}
A standard GNN whose message passing is gated by the causal mask $s$, tying the
prediction to the same edges the explanation reports.

\paragraph{Invariance training.}
The mask is trained with an invariant-causal objective: artificial environments
perturb non-causal context, and an IRM-style penalty forces $s$ to be predictive
\emph{invariantly}, so the top-$k$ edges track the causal motif rather than
spurious structure. (Sparsity and an acyclicity prior regularize the mask.)

\subsection{The deterministic edge certificate}
Let the clean edges be ordered $s(e_1)\ge\cdots\ge s(e_M)$, so
$\Ek=\{e_1,\dots,e_k\}$. The certificate rests on one structural fact:

\begin{proposition}[Edge-locality]
\label{prop:local}
For any perturbation that adds or deletes edges other than $e$, the score
$s(e)$ is unchanged.
\end{proposition}
\noindent This is immediate: $s(e)$ is a function of $e$'s endpoint features
only. Deletions of other edges leave every clean score and rank unchanged, and a
single added edge can occupy at most one slot of the top-$k$. Three guarantees
follow.

\begin{theorem}[Certified overlap]
\label{thm:overlap}
For any perturbation of at most $B$ edges, the perturbed top-$k$ explanation
$\Ek'$ satisfies $|\Ek\cap\Ek'|\ \ge\ k-B$.
\end{theorem}

\begin{theorem}[Per-edge certified radius]
\label{thm:radius}
The rank-$r$ edge of $\Ek$ remains in the top-$k$ explanation under any
perturbation of at most $R(e_r)=k-r$ edges that does not delete it.
\end{theorem}

\begin{theorem}[Transductive hijack impossibility]
\label{thm:hijack}
In the transductive setting, scores of edges outside the training adjacency are
multiplied by a cap $c<1$. If every edge in $\Ek$ has clean score $>c$, then
\emph{no} inserted edge can enter the top-$k$ explanation, for \emph{any} budget
$B$.
\end{theorem}

\noindent Proofs are immediate from Prop.~\ref{prop:local} (counting argument)
and given in the appendix. The certificate is deterministic and closed-form:
unlike voting, there is nothing to sample. We verify empirically that it is
\emph{sound and tight}---a worst-case adversary achieves exactly the bound of
Thm.~\ref{thm:overlap} (Fig.~\ref{fig:tight}).

\paragraph{Why this matters.}
The $\bigO(T)$ sub-graph evaluations of voting certificates are replaced by a
single pass ($\bigO(1)$); the randomized, conservative lower bound is replaced by
an exact one; and, because the scorer is causal, the certified object is a
\emph{faithful} explanation.

% =====================================================================
\section{Experiments}
\label{sec:exp}
\input{experiments}

% =====================================================================
\section{Conclusion}
\label{sec:concl}
We argued that certifying the \emph{stability} of a GNN explanation is
insufficient---a stable explanation can be spurious---and introduced \aeth{},
which certifies the robustness of \emph{causally faithful} explanations with a
deterministic, tight, single-pass certificate. The certificate follows from a
propagation-free causal scorer whose edge-locality we exploit in closed form. On
the state of the art's own metric and published numbers \aeth{} certifies
markedly larger perturbation sizes on real data, remains faithful under spurious
correlation, and certifies at $\bigO(1)$ cost. We hope this reframes certified
explainability around \emph{what} is certified, not only \emph{how robustly}.

% =====================================================================
\bibliographystyle{aaai2026}   % <-- official AAAI .bst goes alongside the .sty
\bibliography{refs}

% =====================================================================
\appendix
\input{appendix}

\end{document}
